---
title: "  "
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)

library(deSolve)
```
## Linearization of non-linear models: SIR models
As we saw previously, SIR models are simple epidemiological compartment models that predict the spread of an infectious disease. In this example, we will use the SIR model as a way of demonstrating how to linearize a non-linear model.

#### SIR model diagram
```{r}
DiagrammeR::grViz("digraph {

# Initialize graph
graph [layout = dot, rankdir = LR]

# Define the nodes and global styles of the nodes. We can override these in box if we wish
node [shape = circle, style = filled]

#Edges in graphs must be between nodes, therefore dummy nodes are needed for signaling edges and input/output edges

OS [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
S [fillcolor = green]
SI [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
I [fillcolor = red]
IO [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
R [fillcolor = blue]

# edge definitions with the node IDs
edge[fontsize = 10]
# Edge weights are not necessary, but can be used to get edges in straight line (higher weight edges have more emphasis)

S -> SI [label = 'rI', weight = 2, dir = none]
SI -> I [weight = 2]
I -> SI [label = '+', tailport = n, style = dashed, fontcolor = blue, color = blue]
OS -> S [label = 'rBirth', weight = 2]
I -> IO [label = 'rDeath', weight = 2]

I -> R [label = 'rR', weight = 1]
R -> S [label = 'rS', weight = 1]
}")
```

#### SIR model definition
```{r}
SIR <- function(t, state, parameters) {# SIR is the name of this function with inputs t, state, parameters 
  # with function executes lines in {} (differential equations) using values defined in state and parameters
  with(as.list(c(state, parameters)),{  
      dS <- rBirth + rS*R - rI*S*I #ODE for susceptible population
      dI <- rI*S*I - rR*I - rDeath*I #ODE for infected population
      dR <- rR*I - rS*R #ODE for recovered population
      list(c(dS, dI, dR)) #Return change in variables
    })
  }

```

#### SIR model parameters
```{r}
parameters <- c(rBirth = 3, #Birth rate
                rDeath = 0.02, #Death rate
                rI = 0.05, #Infection rate
                rR = 0.05, #Recovery rate
                rS = 0.01 #Loss of immunity rate
                )
```

#### SIR initial conditions
Initial conditions in this example are based on slighty perturbing the system from a steady state.
```{r}
ss <- c(S=140,I=150,R=750) # Steady state values of S,I, and R
perturb <- c(dS=30,dI=-30,dR=-50) # Perturbation from SS
state <- with(as.list(c(ss, perturb)),{
            c(S = S + dS, #Initial number of susceptible individuals  
              I = I + dI, #Initial number of infected individuals
              R = R + dR  #Initial number of recovered individuals
            )
})
```

#### SIR model simulation
```{r}

times <- seq(from = 0,to = 300, by = 1) #Definition of time (initial,final,step size)

#ode is like ode45 in Matlab and is used to simulate systems of ODEs
out <- ode(y = state,#Initial conditions 
           times = times, #Time definition
           func = SIR, #Name of function defining ODEs
           parms = parameters #Parameter values
           )
```

#### Plot simulation results
```{r}
plot(out[,"time"],out[,"S"],col=3,lwd=2,type="l",xlab="Time",ylab="Population size",ylim=c(0,800)) #Plot S population vs time
lines(out[,"time"],out[,"I"],col=2,lwd=2,type="l") #Plot I population vs time
lines(out[,"time"],out[,"R"],col=4,lwd=2,type="l") #Plot R population vs time
legend("topright",legend = c("S","I","R"),col = c(3,2,4),lwd = 2) #Define plot legend

```

#### Linearized SIR model definition
To generate the linearized SIR model we simply take the partial derivative of each equation, and evaluate the resulting expression at operating point (steady state). *Note: Not all equations will need to be dependent on all variables.*
```{r}
lSIR <- function(t, state, parameters) {# SIR is the name of this function with inputs t, state, parameters 
  # with function executes lines in {} (differential equations) using values defined in state and parameters
  with(as.list(c(state, parameters)),{  
      dS <- rSR*R + rSS*S + rSI*I #Linearized ODE for susceptible population
      dI <- rII*I +rIS*S #Linearized ODE for infected population
      dR <- rRI*I + rRR*R #Linearized ODE for recovered population
      list(c(dS, dI, dR)) #Return change in variables
    })
  }

```

#### Linearized SIR model parameters
```{r}
Lparameters <- with(as.list(c(ss, parameters)),{ #Parameters for the linearized model are dependent on the 
                                                 # parameter values of the nonlinear model and the SS
                        c(rSS = -rI*I, 
                          rSI = -rI*S, 
                          rSR = rS,
                          rIS = rI*I, 
                          rII = rI*S-rR-rDeath,
                          rRI = rR,
                          rRR = -rS
                          )
})
```

#### Linearized SIR initial conditions
The linearized model is capturing the perturbation from a steady state, therefore, the initial conditions for the linearized model is just the perturbations. *Note: In the linearized model (0,0,0) is a SS.*
```{r}
Lstate <- with(as.list(c(perturb)),{
            c(S = dS, #Initial number of susceptible individuals  
              I = dI, #Initial number of infected individuals
              R = dR  #Initial number of recovered individuals
            )
})
```

#### SIR model simulation
```{r}
times <- seq(from = 0,to = 300, by = 1) #Definition of time (initial,final,step size)

#ode is like ode45 in Matlab and is used to simulate systems of ODEs
Lout <- ode(y = Lstate,#Initial conditions 
           times = times, #Time definition
           func = lSIR, #Name of function defining ODEs
           parms = Lparameters #Parameter values
           )
```

#### Plot simulation results
To transform the linearized model back to the original space the SS values are added to the variable values.
```{r}
plot(Lout[,"time"],Lout[,"S"]+ss["S"],col=3,lwd=2,type="l",xlab="Time",ylab="Population size",ylim=c(0,800)) #Plot S population vs time
lines(Lout[,"time"],Lout[,"I"]+ss["I"],col=2,lwd=2,type="l") #Plot I population vs time
lines(Lout[,"time"],Lout[,"R"]+ss["R"],col=4,lwd=2,type="l") #Plot R population vs time
legend("topright",legend = c("S","I","R"),col = c(3,2,4),lwd = 2) #Define plot legend

```
