---
title: ""
output: html_document
params:
  input: NA
  output: NA
---

```{r setup, include=FALSE}
library(knitr)
library(shiny)
library(DiagrammeR)

knitr::opts_chunk$set(echo = TRUE)
```
## Simplified MAPK Cascade Model
The following report summarizes simulation results from the simplified MAPK cascade model performed on *`r date()`*.

### Model Diagram
```{r echo=F}
grViz("digraph {

                      # Initialize graph
                      graph [layout = dot, rankdir = LR]
                      
                      # Define the nodes and global styles of the nodes. We can override these in box if we wish
                      node [shape = circle, style = filled, fixedsize = true,fontcolor=white]
                      
                      #Edges in graphs must be between nodes, therefore dummy nodes are needed for signaling edges and input/output edges
                      
                      S [fillcolor = dimgray]

                      E [fillcolor = blue]
                      X [fillcolor = blue]
                      fEX [style=invis, shape=point, width=0] #Dummy node
                      rEX [style=invis, shape=point, width=0] #Dummy node
                      XE [fillcolor = blue]
                      iXP [style=invis, shape=point, width=0] #Dummy node
                      XP [fillcolor = blue]
                      
                      Y [fillcolor = red]
                      fYXP [style=invis, shape=point, width=0] #Dummy node
                      rYXP [style=invis, shape=point, width=0] #Dummy node
                      YXP [fillcolor = red]
                      iYP [style=invis, shape=point, width=0] #Dummy node
                      YP [fillcolor = red]

                      Z [fillcolor = ForestGreen]
                      fZYP [style=invis, shape=point, width=0] #Dummy node
                      rZYP [style=invis, shape=point, width=0] #Dummy node
                      ZYP [fillcolor = ForestGreen]
                      iZP [style=invis, shape=point, width=0] #Dummy node
                      ZP [fillcolor = ForestGreen]
          

                      # edge definitions with the node IDs
                      edge[fontsize = 10]
                      # Edge weights are not necessary, but can be used to get edges in straight line 
                      # (higher weight edges have more emphasis)
                      
                      ranksep=0.2;
                      
                      S -> fEX [label = '+', tailport = n, style = dashed, 
                                fontcolor = green, color = green,weight=0]

                      E -> fEX [dir = none, weight=1]
                      X -> fEX [dir = none, weight=2]
                      fEX -> XE [label = 'k@_{on}', weight=2]
                      rEX -> X [weight = 2]
                      rEX -> E [weight = 1]
                      XE -> rEX [weight = 2, dir = none, label = 'k@_{off}']
                      XE -> iXP [label = 'k@_{cat}', weight = 2, dir = none]
                      iXP -> XP [weight = 2]
                      iXP -> E [weight = 1]
                      XP -> X [label = 'k@_{dp}',weight = 1]
                      
                      XP -> fYXP [dir = none, weight=1]
                      Y -> fYXP [dir = none, weight=2]
                      fYXP -> YXP [label = 'k@_{on}', weight=2]
                      rYXP -> Y [weight = 2]
                      rYXP -> XP [weight = 1]
                      YXP -> rYXP [weight = 2, dir = none, label = 'k@_{off}']
                      YXP -> iYP [label = 'k@_{cat}', weight = 2, dir = none]
                      iYP -> YP [weight = 2]
                      iYP -> XP [weight = 1]
                      YP -> Y [label = 'k@_{dp}',weight = 1]

                      YP -> fZYP [dir = none, weight=1]
                      Z -> fZYP [dir = none, weight=2]
                      fZYP -> ZYP [label = 'k@_{on}', weight=2]
                      rZYP -> Z [weight = 2]
                      rZYP -> YP [weight = 1]
                      ZYP -> rZYP [weight = 2, dir = none, label = 'k@_{off}']
                      ZYP -> iZP [label = 'k@_{cat}', weight = 2, dir = none]
                      iZP -> ZP [weight = 2]
                      iZP -> YP [weight = 1]
                      ZP -> Z [label = 'k@_{dp}',weight = 1]
                  }")
```

### Model Parameters
The following tables include the values of the model inputs that were used to generate the results
summarized by this report.

```{r echo=FALSE}
pTab <- data.frame(Concentration=c(params$input$kon,params$input$koff,params$input$kcat,params$input$kdp))
colnames(pTab) <- c("Value")
rownames(pTab) <- c("Association rate (kon)","Dissociation rate (koff)","Catalytic rate (kcat)","Dephosphorylation rate (kdp)")
kable(pTab,caption = "MAPK Enzyme Kinetics Parameters")
```

```{r echo=FALSE}
sTab <- data.frame(Value=c(params$input$A,params$input$freq,params$input$startS,params$input$endS))
colnames(sTab) <- c("Value")
rownames(sTab) <- c("Signal amplitude (A)","Signal frequence (freq)",
                    "Signal start time (startS)","Signal end time (endS)")
kable(sTab,caption = "MAPK Signal Parameters")
```

### Plots Summarizing the Simulation
&nbsp;

```{r echo=FALSE}
out <- params$output # Call reactive function to get updated simulation 
    
# Plots results of simulation and is ran everytime the input is updated
par(mar=c(4,4,1,1))
plot(out[,1],out[,5],ylim=c(0,ceiling(max(out[,c(5,8,11)]))),xlab="Time",ylab="Population",col=4,lwd=2,type="l",cex.lab=1.3,cex.axis=1.3)
lines(out[,1],out[,8],xlab="Time",ylab="",col=2,lwd=2,type="l")
lines(out[,1],out[,11],xlab="Time",ylab="",col=3,lwd=2,type="l")
legend("topright",c("XP","YP","ZP"),lwd = 2, col=c(4,2,3))
```
&nbsp;
&nbsp;

**Figure 1.** *Time course simulation of MAPK cascade activation.*  
&nbsp;
&nbsp;

