---
title: "  "
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)

library(DiagrammeR)
```
## Markov Chain Models
Markov models describe a system that can assume m different states. At each (discrete) time point, the system is in exactly one of these states. In discrete time steps, the system transitions with some probability into a 	different state or stays where it is. 

Example: Pinball machine with flippers, bumpers, kickers, slingshots, …

#### Define Markov Model transition probabilities
```{r}

#c combines numbers/characters into a vector 
#rbind builds matrices by combining vectors row-wise
MM <- rbind(c(0.79,0.19,0.02), #Probability of being in state 1 and moving to state 1,2,3
             c(0.19,0.54,0.27), #Probability of being in state 2 and moving to state 1,2,3
             c(0.51,0.43,0.06) #Probability of being in state 3 and moving to state 1,2,3
             )
MM <- as.matrix(MM) #Make MM have matrix object type
rownames(MM) <- 1:nrow(MM) #Add row names
colnames(MM) <- 1:nrow(MM) #Add col names

graph <- # Visualize system with edge weights indicating probability of transition
  create_graph() %>%
  add_full_graph(n = nrow(MM), 
                 keep_loops = TRUE,
                 type = "weighted",
                 edge_wt_matrix = t(MM)
) %>%
  copy_edge_attrs(
    edge_attr_from = weight,
    edge_attr_to = label
    )

render_graph(graph) #function from  DiagrammeR visualizing a graph
```

#### Markov Matrix analysis
```{r}
iPos <- 1 # Initial position
nTrans <- 1 # Number of transitions

ps <- ((1:nrow(MM))==iPos)+0 #Initialize probability distribution
ps <- t(t(ps)) #Make initial prob a column vector

for(i in 1:nTrans){
  ps <- t(MM)%*%ps # Update probability distribution for each step/transition
}

ps
```
