---
title: "Introduction to Modeling Biological Systems"
output: learnr::tutorial
runtime: shiny_prerendered
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(learnr)
library(DiagrammeR)

knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## Introduction

In this tutorial we will cover the following modeling tasks:

1.  How to generate system diagrams.
2.  How to perform dynamic simulations of ODE-based models.

### Prerequisite

The tasks in this tutorial require the following packages be installed:

1.  `DiagrammeR` - package for generating graphs and diagrams.
2.  `deSolve` - package for simulating systems of ordinary differential
    equations.\

### R Tip: *How to install packages in R*

You can use `install.packages('package')` to install any necessary
package.

```{r install_packages, echo = FALSE}
question("Which code installs `DiagrammeR` package successfully? ",
         answer("`install.packages(DiagrammeR)`", message = "Quotation marks are missing."),
         answer("`install.packages('diagrammer')`", message = "R package names are case sensitive."),
         answer("`install.packages('DiagrammeR')`", correct = T), 
         allow_retry = T)
```

## DiagrammeR

For generating diagrams, such as though used in visualizing biological
systems, `DiagrammeR` supports Graphviz. From the DiagrammeR
documentation:

- Graphviz consists of a graph description language called the DOT
  language and it also comprises various tools that can process the
  DOT language. DOT is highly customizable and it allows you to
  control line colors, arrow shapes, node shapes, and many other
  layout features.
  
- The Graphviz graph specification must begin with a directive stating 
  whether a directed graph (digraph) or an undirected graph (graph) 
  is desired. Semantically, this indicates whether or not there is a natural 
  direction from one of the edge’s nodes to the other. An optional graph ID 
  follows this and paired curly braces denotes the body of the statement list. 

- Here is the basic structure: `[strict] (graph | digraph) [ID] '{' stmt_list '}'`

- The graph statement, the node statement, and the edge statement are the three most 
  commonly used statements in the Graphviz DOT language. Graph statements allow for 
  attributes to be set for all components of the graph. Node statements define and 
  provide attributes for graph nodes. Edge statements specify the edge operations 
  between nodes and they supply attributes to the edges. In the examples that follow
  will be using directed graphs, therefore an edge must be specified using the edge 
  operator `->`.
  
- Within these statements follow statement lists. Thus for a node statement, a list 
  of nodes is expected. For an edge statement, a list of edge operations. Any of the 
  list items can optionally have an attribute list (attr_list) which modify the attributes 
  of either the node or edge.
  
More information can be found here: (https://rich-iannone.github.io/DiagrammeR/articles/graphviz-mermaid.html)

```{r grViz, echo = FALSE}
question("Which psuedocode is correct for using Graphviz for generating SIR graphs?",
         answer("`DiagrammeR::grViz('digraph{body}')`", correct = T),
         answer("`DiagrammeR::Graphicviz('digraph{body}')`", message = "`grViz()` function renders graphs"),
         answer("`DiagrammeR::grViz('graph{body}')`", message = "SIR graphs are directed"), 
         allow_retry = T)
```

## Example Graphviz Diagram 
To start a Graphviz diagram, we will use the `grViz()` function, and pass to it
a Graphviz statement defining the graph. To initialize a directed graph we will use `digraph` 
and then define the nodes and edges and specifying the styles, such as the shape of 
the of nodes etc. See the following link for more details regarding attributes and stules: <https://rich-iannone.github.io/DiagrammeR/articles/graphviz-mermaid.html#graphviz-attributes>. 
Suggested steps are as follow:

1. Write graph statement.
2. Write node statement(s).
3. Write edge statement(s).

Here is an example where nodes (in this case styled as boxes and circles) can be easily defined along with their connections: 
```{r}
grViz("digraph boxes_and_circles {
  # a 'graph' statement
  graph [overlap = true, fontsize = 10]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  A; B; C; D; E; F

  node [shape = circle,
        fixedsize = true,
        width = 0.9] // sets as circles
  1; 2; 3; 4; 5; 6; 7; 8

  # several 'edge' statements
  A->1 B->2 B->3 B->4 C->A
  1->D E->A 2->4 1->5 1->F
  E->6 4->6 5->7 6->7 3->8

}")

```

## Diagram example: SIR models
To introduce the concepts of generating a model diagram and visualizing it 
using `DiagrammeR`, we will use the SIR models as the motivating example. 
SIR models are simple epidemiological compartment models that predict
the spread of an infectious disease. In addition to an infected
population (I), the model also includes a recovered (R), or at least
a temporarily immune population, as well as, a susceptible population (S)
of people who can catch the disease from an infected individual. These
simple models can be powerful in exploring the effects of the various
aspects (rates) of the spread of an infection on the size of the
infected population. 

In the diagram, each population is represented by a "bubble" (node) and each 
process is represented by an arrow (directed edge) indicating flows of people
in/out or between the three populations. Each edge in this example is labeled 
based on the associated rate: birth rate (rB), infection rate (rI), death rate 
(rD), recovery rate (rR), and susceptibility rate (rS).

In a series of excercises that follows, we will walk through how do generate the
diagram below using `DiagrammeR`.

```{r SIR, echo = FALSE, eval=TRUE}
grViz("digraph {

# Graph statement
graph [layout = dot, rankdir = TB]

# Define the global styles of the nodes. 
node [shape = circle, style = filled]

# Node statements and node specific styles.  
S [fillcolor = green]
I [fillcolor = red]
R [fillcolor = blue]

# Edges in graphs must be between nodes, therefore dummy nodes are needed for 
# signaling edges and input/output edges
OS [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
SI [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
IO [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node

# Define global edge properties
edge[fontsize = 10]

# Define flow edges and label wiith rates
OS -> S [label = 'rB']
S -> SI [label = 'rI', dir = none]
SI -> I
I -> IO [label = 'rD']
I -> R [label = 'rR']
R -> S [label = 'rS']

# Define signal edge
I -> SI [label = '+', tailport = n, style = dashed, fontcolor = blue, color = blue]

# Set rank of nodes equal to align
{rank = same; OS; S; I; SI; IO} 

}")
```

## SIR diagram: Graph and node statements
To initilize the graph, the layout of the graph and the direction should be defined 
in the a graph statement using the syntax `graph [layout, rankdir]`. For this example, 
we want to generate a graph with `layout = dot` and `rankdir = TB`. Next, we need to
define the global styles of the nodes using a node statement: `node [attributes]`.
The SIR diagram uses `shape = circle` and `style = filled`. Subsequently, we need to
define the nodes S, I, and R using fillcolor to make S, I, and R, green, red, and blue,
respectively (i.e. `S [fillcolor = green]`).

```{r SIRinit, exercise = TRUE, exercise.lines = 15}
grViz("digraph{
# Graph statement


# Define the global styles of the nodes. 


# Node statements and node specific styles.  


}")
```

```{r SIRinit-solution}
grViz("digraph{
# Graph statement
graph [layout = dot, rankdir = TB]

# Define the global styles of the nodes. 
node [shape = circle, style = filled]

# Node statements and node specific styles.  
S [fillcolor = green]
I [fillcolor = red]
R [fillcolor = blue]

}")
```

## SIR diagram: Dummy nodes
In `Graphviz`/`DiagrammeR` edges can only be added to connect two nodes, but in 
systems biology diagrams we need to be able describe processes that either only
enter or exit a node with a origin/destination node. Also, we often want represent 
how a node can affect a process/flow, such as the signalling edge that accounts 
for the affect of the size of the infected population on the rate of infection.
To be able to account for these types of relationships we need to define "dummy"
nodes that serve as a place holder for these connections. In the SIR example we 
need a dummy node for the influx into S, called OS, which we define as follows:\

`OS [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node`\

Use this as a template to add the OS node, plus a similar edge for the outflux 
from I, IO, and a node for the signaling edge between S and I, SI. 

```{r SIRdummy, exercise = TRUE, exercise.eval = TRUE, exercise.lines = 20}
grViz("digraph {
# Graph statement
graph [layout = dot, rankdir = TB]

# Define the global styles of the nodes. 
node [shape = circle, style = filled]

# Node statements and node specific styles.  
S [fillcolor = green]
I [fillcolor = red]
R [fillcolor = blue]

# Edges in graphs must be between nodes, therefore dummy nodes are needed for 
# signaling edges and input/output edges


}")
```

```{r SIRdummy-solution}
grViz("digraph {

# Graph statement
graph [layout = dot, rankdir = TB]

# Define the global styles of the nodes. 
node [shape = circle, style = filled]

# Node statements and node specific styles.  
S [fillcolor = green]
I [fillcolor = red]
R [fillcolor = blue]

# Edges in graphs must be between nodes, therefore dummy nodes are needed for 
# signaling edges and input/output edges
OS [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
SI [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
IO [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node

}")
```

## SIR diagram: Edge statements and adding flow edges
Next, we need to define the global styles of the edges using an edge statement: 
`edge [attributes]`. The SIR diagram uses `fontsize = 10`. Subsequently, we need to
define the nodes all of the flow edges represented by the same style solid black 
lines. This is where we need to use dummy nodes to account for all of the processes
described earlier. Use the completed diagram below as a guide, and use statements of 
the type: `node1 -> node2 [label]`. The S to I edges are an exception where S to SI 
has a label but also has `dir = none` to avoid having an extra arrow, and SI to I does
not have a label.

```{r SIR, echo = FALSE, eval=TRUE}
```

```{r SIRflow, exercise = TRUE, exercise.eval = TRUE, exercise.lines = 25}
DiagrammeR::grViz("digraph {

# Graph statement
graph [layout = dot, rankdir = TB]

# Define the global styles of the nodes. 
node [shape = circle, style = filled]

# Node statements and node specific styles.  
S [fillcolor = green]
I [fillcolor = red]
R [fillcolor = blue]

# Edges in graphs must be between nodes, therefore dummy nodes are needed for 
# signaling edges and input/output edges
OS [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
SI [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
IO [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node

# Define global edge properties


# Define flow edges and label wiith rates


}")

```

```{r SIRflow-solution}
DiagrammeR::grViz("digraph {

# Graph statement
graph [layout = dot, rankdir = TB]

# Define the global styles of the nodes. 
node [shape = circle, style = filled]

# Node statements and node specific styles.  
S [fillcolor = green]
I [fillcolor = red]
R [fillcolor = blue]

# Edges in graphs must be between nodes, therefore dummy nodes are needed for 
# signaling edges and input/output edges
OS [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
SI [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
IO [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node

# Define global edge properties
edge[fontsize = 10]

# Define flow edges and label wiith rates
OS -> S [label = 'rB']
S -> SI [label = 'rI', dir = none]
SI -> I
I -> IO [label = 'rD']
I -> R [label = 'rR']
R -> S [label = 'rS']

}")

```

## SIR diagram: Finishing touches
To finalize the diagram, all that remains is to add the signaling edge between
node I and dummy node SI. To style the edge as in the example diagram we need to 
set tailport = n, style = dashed, fontcolor = blue, and color = blue. Lastly, to 
get the diagram oriented left to right as in the example we need to set the rank of
the nodes we want to be aligned to be the same, in this case OS, S, I, SI, and IO, 
by adding the following statement at the end `{rank = same; OS; S; I; SI; IO} `.

```{r SIR, echo = FALSE, eval=TRUE}
```

```{r SIRfinal, exercise = TRUE, exercise.eval = TRUE, exercise.lines = 30}
DiagrammeR::grViz("digraph {

# Graph statement
graph [layout = dot, rankdir = TB]

# Define the global styles of the nodes. 
node [shape = circle, style = filled]

# Node statements and node specific styles.  
S [fillcolor = green]
I [fillcolor = red]
R [fillcolor = blue]

# Edges in graphs must be between nodes, therefore dummy nodes are needed for 
# signaling edges and input/output edges
OS [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
SI [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
IO [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node

# Define global edge properties
edge[fontsize = 10]

# Define flow edges and label wiith rates
OS -> S [label = 'rB']
S -> SI [label = 'rI', dir = none]
SI -> I
I -> IO [label = 'rD']
I -> R [label = 'rR']
R -> S [label = 'rS']

# Define signal edge


# Set rank of nodes equal to align

}")
```

```{r SIRfinal-solution}
DiagrammeR::grViz("digraph {

# Graph statement
graph [layout = dot, rankdir = TB]

# Define the global styles of the nodes. 
node [shape = circle, style = filled]

# Node statements and node specific styles.  
S [fillcolor = green]
I [fillcolor = red]
R [fillcolor = blue]

# Edges in graphs must be between nodes, therefore dummy nodes are needed for 
# signaling edges and input/output edges
OS [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
SI [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
IO [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node

# Define global edge properties
edge[fontsize = 10]

# Define flow edges and label wiith rates
OS -> S [label = 'rB']
S -> SI [label = 'rI', dir = none]
SI -> I
I -> IO [label = 'rD']
I -> R [label = 'rR']
R -> S [label = 'rS']

# Define signal edge
I -> SI [label = '+', tailport = n, style = dashed, fontcolor = blue, color = blue]

# Set rank of nodes equal to align
{rank = same; OS; S; I; SI; IO} 

}")
```

## Exercise 1: SIR with quarantine
Your task is to generate a diagram for a modified SIR model that accounts for quarantine,
where some portion of the infected population enters quarantined population and eventually 
enters the recovered population. Starting from the SIR diagram that we just completed, add
the necessary node(s) and edge(s) to account for quarantine. 


```{r SIRQ, exercise = TRUE, exercise.eval = TRUE}
DiagrammeR::grViz("digraph {

# Graph statement
graph [layout = dot, rankdir = TB]

# Define the global styles of the nodes. 
node [shape = circle, style = filled]

# Node statements and node specific styles.  
S [fillcolor = green]
I [fillcolor = red]
R [fillcolor = blue]

# Edges in graphs must be between nodes, therefore dummy nodes are needed for 
# signaling edges and input/output edges
OS [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
SI [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
IO [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node

# Define global edge properties
edge[fontsize = 10]

# Define flow edges and label wiith rates
OS -> S [label = 'rB']
S -> SI [label = 'rI', dir = none]
SI -> I
I -> IO [label = 'rD']
I -> R [label = 'rR']
R -> S [label = 'rS']

# Define signal edge
I -> SI [label = '+', tailport = n, style = dashed, fontcolor = blue, color = blue]

# Set rank of nodes equal to align
{rank = same; OS; S; I; SI; IO} 


}")
```

```{r SIRQ-solution}
DiagrammeR::grViz("digraph {

# Graph statement
graph [layout = dot, rankdir = TB]

# Define the global styles of the nodes. 
node [shape = circle, style = filled]

# Node statements and node specific styles.  
S [fillcolor = green]
I [fillcolor = red]
R [fillcolor = blue]
Q [fillcolor = white]

# Edges in graphs must be between nodes, therefore dummy nodes are needed for 
# signaling edges and input/output edges
OS [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
SI [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node
IO [fixedsize = TRUE, width = 0, height = 0, label = ''] #Dummy node

# Define global edge properties
edge[fontsize = 10]

# Define flow edges and label wiith rates
OS -> S [label = 'rB']
S -> SI [label = 'rI', dir = none]
SI -> I
I -> IO [label = 'rD']
I -> R [label = 'rR']
R -> S [label = 'rS']
I -> Q [label = 'rQ']
Q -> R [label = 'rRQ']

# Define signal edge
I -> SI [label = '+', tailport = n, style = dashed, fontcolor = blue, color = blue]

# Set rank of nodes equal to align
{rank = same; OS; S; I; SI; IO} 

}")
```

## SIR models: Model equations

Now that we have sketched our system in diagram form, a typical next step would be 
to simulate the model to understand the effect of `parameters` (process/flow rates) 
on different the `state` (initial population) by defining/solving ordinary differential
equations (ODEs), like those shown below:

```{r SIRodes, echo=F}
mainPanel(br(), width=350,
                 "$$dS = rBirth + rS \\cdot  R - rI \\cdot  S \\cdot  I $$",
                 "$$dI = rI \\cdot  S \\cdot  I - rR \\cdot  I - rDeath \\cdot  I$$",
                 "$$dR = rR \\cdot  I - rS \\cdot  R$$"
               )
```

## SIR models: Simulating dynamics
For generating a SIR model, we need to define the connection between the parameters 
mentioned above, and to fulfill that functions need to be constructed. The
psuedocode for writing a function in R is as follow:

`function name` \<- `fucntion(independent variables)` {

`body of the function`

}

For analyzing SIR model, it is better to group different `state` and
`parameters` together, and define a function and set `state` and
`parameters` as independent variables. One can define any variables as
below:

```{r sir_model5, exercise = FALSE, exercise.eval = FALSE}
parameters <- c(rBirth = 3, #Birth rate
                rDeath = 0.02, #Death rate
                rI = 0.0005, #Infection rate
                rR = 0.05, #Recovery rate
                rS = 0.01 #Loss of immunity rate
                )

state <- c(S = 1000, #Initial number of susceptible individuals`  
           I = 10, #Initial number of infected individuals
           R = 0 #Initial number of recovered individuals
          )
```

Then, using ODE equations, we can define SIR model as a function of
`state`, `parameters` as below which generates change in `state` in a
`list` format:

```{r sir_model6, exercise = F, exercise.eval = FALSE}
SIR <- function(state, parameters) {# SIR is the name of this function with inputs t, state, parameters 
  # with function executes lines in {} (differential equations) using values defined in state and parameters
  with(as.list(c(state, parameters)),{  
      dS <- rBirth + rS*R - rI*S*I #ODE for susceptible population
      dI <- rI*S*I - rR*I - rDeath*I #ODE for infected population
      dR <- rR*I - rS*R #ODE for recovered population
      list(c(dS, dI, dR)) #Return change in variables
    })
  }
```

It is important to note that without the `with` function, the function
wouldn't work as it doesn't have access to the outside variables `state`
and `parameters`.

For solving the ODE equations, `deSolve` library needs be installed.
Using `ode` function and by defining `y` (initial condition), `times`
(time definition which can be defined using `seq(start, to, interval)`
function), `func` (functions containing ODEs), and `parms` (parameters),
ODEs can be solved. `ode` is like ode45 in Matlab and is used to
simulate systems of ODEs. Note that the function that contains the ODEs
now needs to have `t` (time) as a variable when defining. Then:

`output <- ode(times, y, func, parms)`

```{r sir_model8, echo = FALSE}
question("For solving ODE equations of SIR model using variables defined previously, which one 
         is corrrect?",
         answer("`y = states`, `times = seq(from = 0,to = 100, by = 1)`, `func = SIR(t, state, parameters)`, `parms = parameters`", correct = T),
         answer("`y = parameters`, `times = seq(from = 0,to = 100, by = 1)`, `func = SIR(t, state, parameters)`, `parms = state`", message = "`y` would be the `state`"),
         answer("`y = states`, `times = seq(from = 0,to = 100, by = 1)`, `func = SIR(state, parameters)`, `parms = parameters`", message ="SIR should containe `t` as an independent variable")) 
         
```

Complete the following code by defining the an ODE function `SIR` and
set `t`, `states`, and `parameters` as its variables. Then, simulate
using `ode` function for 100 days. Make sure to call `deSolve` library
before using `ode` function.

```{r sir_model9, exercise = TRUE}
parameters <- c(rBirth = 3, #Birth rate
                rDeath = 0.02, #Death rate
                rI = 0.0005, #Infection rate
                rR = 0.05, #Recovery rate
                rS = 0.01 #Loss of immunity rate
                )
state <- c(S = 1000, #Initial number of susceptible individuals`  
           I = 10, #Initial number of infected individuals
           R = 0 #Initial number of recovered individuals
          )
```

```{r sir_model9-solution}

SIR <- function(t, state, parameters) {# SIR is the name of this function with inputs t, state, parameters 
  # with function executes lines in {} (differential equations) using values defined in state and parameters
  with(as.list(c(state, parameters)),{  
      dS <- rBirth + rS*R - rI*S*I #ODE for susceptible population
      dI <- rI*S*I - rR*I - rDeath*I #ODE for infected population
      dR <- rR*I - rS*R #ODE for recovered population
      list(c(dS, dI, dR)) #Return change in variables
    })
  }
times <- seq(from = 0,to = 100, by = 1) #Definition of time (initial,final,step size)

#ode is like ode45 in Matlab and is used to simulate systems of ODEs
out <- ode(y = state,#Initial conditions 
           times = times, #Time definition
           func = SIR, #Name of function defining ODEs
           parms = parameters #Parameter values
           )
```

## Plotting in R

As we solved the ODEs and now have information about how different
`state` changes as days go by, it is time to analyze our results from
previous parts. One of the interesting approach to fully understand what
is happening, is by making meaningful plots. In R programming language,
`plot` function is used for generating plots.Optionally, you can change
the style of the plot depending on your application. Colors `col`, line
width `lwd`, axis labels `xlab` and `ylab` and the plot type `type`, and
others. can be edited for a better representation. More information can
be found here
(<https://www.datamentor.io/r-programming/plot-function/>). Then:

`plot(x, y, col, lwd, xlab, ylab)`

You can add more functions to the plot through a chosen plot type `type`
such as `lines`:

`lines(x, y, col, lwt, xlab, ylab)`

And at the end, you can edit the legends:

`legends(location, title, col, lwd)`

You can include a vector to represent multiple plots in `legend`.

For example, you want to make a plot of positive COVID-19 and flue cases
in a small town in 2020 by month.

```{r plot1, exercise = TRUE, exercise.eval = TRUE}
month = c(1,2,3,4,5,6,7,8,9,10,11,12)
covid.positive = c(7, 29, 85, 92, 107, 188, 159, 150, 150, 118, 67, 30)
flue.poistive = c(12, 29, 10, 9, 8, 12, 3, 0, 0, 3, 1, 0)
plot(month, covid.positive, col = 2, type = "l", xlab = "month", ylab = "positive cases") 
lines(month, flue.poistive, col = 3, type = "l")
legend("topleft", legend = c("COVID-19 cases", "Flue cases"), col = c(2,3), lwd = 1)
```

## Slicing and Indexing in R

Indexing and Slicing are operations by which one select desired values
from a vector. In general, this feature of R is very useful, as you
might want to analyze part of your result, and then compare it to
others. For example, you have a vector `A= c(1,2,3,4,5)`, and you want
to extract `c(2,3,4)`, then:

```{r slicing1, exercise = TRUE, exercise.eval = TRUE}
A = c(1,2,3,4,5)
A[2:4]
```

Previously, we generated a data frame containing different `state` (`S`,
`I` and `R`) at different time, and called it `out`. vectors are used to
hold values of the same datatype; however, data frames are 2 dimensional
(rows and columns) objects. If you are only interested in parts of your
result, then:

`out[row(s),column(s)]`

For example, you are only interested in the data from the first four
days, then:

`out[1:5,]`

It is important to note that if you leave the row (or column) blank, R
will select entire rows (columns). Alternatively, you can extract
columns (or rows) by their IDs if they have one. `ode` generates the
results with the corresponding IDs, so you can access data of different
types of population through their IDs.

```{r slicing2, echo = FALSE}
question("You are only interersted in the Recovered population from day five to the end of the day ten. Which one is the correct syntax to extract the recovered population from `out`?",
         answer("`out[4:11 'R']`", message = "`,` is needed."),
         answer("`out[4:11, R]`", message = "R should be in quotaion mark."),
         answer("`out[4:11]`", message = "Recovered population must be selected"), 
         answer("`out[4:11, 'R']`", correct = TRUE), 
         allow_retry = TRUE
)
```

## SIR model: Plotting time-dependent populations
Through `plot` function and slicing, we can now plot the population size
of different `state` versus time.

```{r sirplot, exercise = TRUE, exercise.eval = TRUE}
library("deSolve")
SIR <- function(t, state, parameters) {# SIR is the name of this function with inputs t, state, parameters 
  # with function executes lines in {} (differential equations) using values defined in state and parameters
  with(as.list(c(state, parameters)),{  
      dS <- rBirth + rS*R - rI*S*I #ODE for susceptible population
      dI <- rI*S*I - rR*I - rDeath*I #ODE for infected population
      dR <- rR*I - rS*R #ODE for recovered population
      list(c(dS, dI, dR)) #Return change in variables
    })
}

parameters <- c(rBirth = 3, #Birth rate
                rDeath = 0.02, #Death rate
                rI = 0.0005, #Infection rate
                rR = 0.05, #Recovery rate
                rS = 0.01 #Loss of immunity rate
                )

state <- c(S = 1000, #Initial number of susceptible individuals  
           I = 10, #Initial number of infected individuals
           R = 0 #Initial number of recovered individuals
          )

times <- seq(from = 0,to = 100, by = 1) #Definition of time (initial,final,step size)

#ode is like ode45 in Matlab and is used to simulate systems of ODEs
out <- ode(y = state,#Initial conditions 
           times = times, #Time definition
           func = SIR, #Name of function defining ODEs
           parms = parameters #Parameter values
           )

plot(out[,"time"],out[,"S"],col=3,lwd=2,type="l",xlab="Time",ylab="Population size",ylim=c(0,1000)) #Plot S population vs time
lines(out[,"time"],out[,"I"],col=2,lwd=2,type="l") #Plot I population vs time
lines(out[,"time"],out[,"R"],col=4,lwd=2,type="l") #Plot R population vs time
legend("topright",legend = c("S","I","R"),col = c(3,2,4),lwd = 2) #Define plot legend

```

## Exercise 2: Simulate SIRQ model

Simulate SIRQ model for 100 days with following variables:
`parameters <- c(rBirth = 3, rDeath = 0.02, rI = 0.0005, rR = 0.05, rS = 0.01, rQ = 0.1, rRI = 0.01)`
`state <- c(S = 1000,  I = 10, R = 0, Q = 0)`
and plot the population size of different `state` versus time.
```{r sirq1, exercise = TRUE}

```

```{r sirq1-solution}

SIRQ <- function(t, state, parameters) {
    with(as.list(c(state, parameters)), {
      dS <- rBirth + rS*R - rI*S*I + rRI*Q
      dI <- rI*S*I - rR*I - rDeath*I - rQ*I
      dR <- rR*I - rS*R
      dQ <- rQ*I - rRI*Q
      list(c(dS, dI, dR, dQ))
    })
}

parameters <- c(rBirth = 3, 
                rDeath = 0.02,
                rI = 0.0005, 
                rR = 0.05, 
                rS = 0.01,
                rQ = 0.1, 
                rRI = 0.01
                )

state <- c(S = 1000,  
           I = 10, 
           R = 0,
           Q = 0
          )

times <- seq(from = 0,to = 100, by = 1)

out <- ode(y = state, 
           times = times, 
           func = SIRQ, 
           parms = parameters
           )

plot(out[,"time"],out[,"S"],col=3,lwd=2,type="l",xlab="Time",ylab="Population size",ylim=c(0,1000))
lines(out[,"time"],out[,"I"],col=2,lwd=2,type="l")
lines(out[,"time"],out[,"R"],col=4,lwd=2,type="l")
lines(out[,"time"],out[,"Q"],col=1,lwd=2,type="l")
legend("topright",legend = c("S","I","R","Q"),col = c(3,2,4,1),lwd = 2)
```
